Archaic introgression and the distribution of shared functional variation under stabilizing selection

### 2023-09-27 doc/manuscript.tex
update figure
edits to intro

@@

@@

\title{Stabilizing selection, archaic introgression, and the distribution
of shared functional variation}
\author{APR}
\date{\today}

@@

@@

Genomic studies have shown that admixture is a common occurrance between
diverged populations and closely related taxa \citep{brandvain2014speciation,
skoglund2015ancient,suvorov2022widespread}, and potentially has been widespread
in primate \citep{tung2017contribution,sorensen2023genome} and hominin
\citep{wolf2018outstanding,peter2020100} evolution. Understanding the effects
of admixture on phenotypic and molecular variation is therefore relevant to
many natural systems, including to studies of the genetic basis of complex
traits and diseases. In particular, archaic introgression in \emph{Homo
sapiens} involving Neanderthals and Denisovans has recently attracted
considerable attention, both in understanding the historical processes leading
to observed distributions of introgressed DNA in present-day populations
\citep{prufer2014complete,villanea2019multiple,chen2020identifying} as well as
its contribution to quantitative traits
\citep{sankararaman2016combined,wei2023lingering}.

After their introduction through admixture, introgressed alleles may be
selected for or against. Some introgressed haplotypes appear to have been
positively selected in modern humans
\citep{huerta2014altitude,racimo2017signatures,enard2018evidence,gower2021detecting},
possibly because such alleles are locally adaptive and provide a fitness
advantage when encountering a novel environment. Despite these cases of
adaptive introgression, introgressed alleles were more likely to have been
negatively selected in modern humans
\citep{harris2016genetic,juric2016strength}. As Neanderthal and Denisovan
population sizes are thought to have been relatively small for hundreds of
thousands of years, theory predicts that they would accumulate deleterious
variation at a faster rate than larger populations. Introgressed haplotypes
would then carry more deleterious variants, which would quickly be selected
against after admixture. \aprcomment{Double check the following statements.} In
mapping the distribution of haplotypes identified as introgressed from
Neanderthals, there is a reduction of Neanderthal-related ancestry in enhancers
and regulatory regions
\citep{petr2019limits,telis2020selection,yermakovich2023long}. This
Neanderthal-allelic depletion, or ``deserts,'' support the hypothesis that
introgressed functional alleles were selected against
\citep{sankararaman2014genomic,sankararaman2016combined}.

There is genetic evidence that early \emph{H. sapiens} reciprocally contributed
to Neanderthal genomes \citep{kuhlwilm2016ancient,hubisz2020mapping}, with
introgression in the reverse direction occurring tens or hundreds of thousands
of years before the Neanderthal-to-human event during the global dispersal of
later humans. This is supported by ``near modern'' \emph{H.  sapiens} outside
of African around 120--100ka or earlier
\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
with Neanderthals and providing opportunities for early contacts. While
estimates of the genomic contribution of early \emph{H. sapiens} to
Neanderthals vary \citep{kuhlwilm2016ancient,hubisz2020mapping}, up to 3--5\%
of later Neanderthal genomes may have been contributed through introgression
from humans to Neanderthals. Under the reasoning that \emph{H.
sapiens}-related haplotypes carried fewer deleterious alleles,
human-introgressed DNA would have been favored in Neanderthal genomes. The
replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports this model \citep{posth2017deeply,petr2020evolutionary}.

Models for selection against introgressed alleles are often based on load
arguments (in particular, differences in the rate of accumulation of
unconditionally deleterious variation in populations of different sizes) or
hybrid incompatibilities \citep{muller1942isolating}. These arguments are
founded primarily in population genetics theory, with little notion of genetic
variation contributing to phenotypic variation that selection operates on.
Most phenotypic traits are thought to be under stabilizing selection
\citep{sanjak2018evidence,sella2019thinking}, including gene regulation
\citep{gilad2006natural,hodgins2015gene,price2022detecting}. This is relevant
to selection on Neanderthal-introduced alleles in modern humans, as the some of
the strongest signals of selection against such alleles occur in regulatory
regions \citep{sankararaman2014genomic}.

Stabilizing selection on a trait acts to maintain the phenotypic distribution
of that trait near some trait optimum, which is achieved by reducing phenotypic
variation (Figure~\ref{fig:one-pop}A). When the mean phenotype of the
population is close to the phenotypic optimum, classical models predict that
alleles contributing to such a trait are subject to underdominant selection,
i.e., selection against the minor allele at a locus
\citep{robertson1956effect}. This has proven to be a useful model for
understanding allele frequencies of variants contributing to a trait under
stabilizing selection in the single-population setting
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Stabilizing selection and additive genetic variance in one population.}
        \aprcomment{See comparisons to simulations assuming free recombination in
        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
    }
    \label{fig:one-pop}
\end{figure}

In two diverged populations, the genomic architecture of a trait under
stabilizing selection has a higher rate of turnover compared to neutrally
evolving loci \citep{yair2022population}. We might therefore expect an
accumulation of fixed differences between the \emph{H. sapiens} and Neanderthal
branches for alleles contributing to a trait, even when the mean phenotype in
each population is close to the same trait optimum. When alleles that are fixed
in one population are introduced to another population in which they were
previously absent, they will be at low frequency and subsequently selected
against. Likewise, if the ancestral allele is reintroduced to a population
fixed for a derived allele, the ancestral allele will be at low frequency and
will be selected against. In either case we should expect the introgressed
allele, whether ancestral or derived, to be selected against, regardless the
historical relative sizes of the populations involved. This prediction
contrasts with the population-genetics ``load'' model, in which haplotypes with
fewer deleterious variants (such as those from the population with larger
historical size) are favored after introgression in either direction.

@@

We consider a polygenic trait such that an individual's (additive) genetic
value is the sum over all effects of alleles in their genome, so that for
individual $i$, \(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size
of the derived allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their
genotype at that locus (i.e., the number of alleles they carry at that locus).
For unlinked loci, the expected additive genetic variance is \(V_A = \sum_l
2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We ignore
dominance and epistatic effects, so that \(V_G = V_A\), assuming no linkage.
We can further ingore environmental effects \citep{simons2018population}, so
that the phenotypic variance \(V_P=V_A\).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $o$, and we assume a Gaussian fitness function so that relative fitness
is given by \(f(G_i | O, V_S) = \exp{(-(G_i - O) / 2 V_S)}\). $V_S$ can be
interpreted as the strength of selection on the trait, where larger $V_S$
implies weaker selection. For a population with mean phenotype at (or very
close to) the optimum, the mean fitness of the population (assuming a roughly
normal distribution of phenotypic values in the population) is
\[\bar{w} \approx \int_{-\infty}^\infty f(G | O, V_S) \mathcal{N}(0, V_G) dG
\approx \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\]
Thus, as the genetic variance increases, mean fitness among individuals in the
population decreases.

@@

If all alleles contribute equally to the trait (as \(\pm a\)),
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_{G,t})\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known result
\[\tilde{V_G} \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
Mutations are not expected to each have the same effect size $a$, but
rather will be drawn from some distribution. Here, we will assume mutation
effect sizes are drawn from a normal distribution with mean 0 and given
variance $V_M$.

@@

We extend the moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017} to include underdominance with given selection
coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles with effect
size $a$ to the total genetic variance $V_G$ is then found by computing the
expected pairwise diversity from the SFS (with sample size $n$, $\Phi_n$), as
\[V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da.\]
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. The total
genetic variance is then 
\[V_G=\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration, and admixture. In this
study, we consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In the simplest model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, on remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{toy-admixture}A,B).

@@

>>>>

\usepackage{setspace}

@@

\DeclareMathOperator{\Var}{Var}

@@

\title{Archaic introgression and the distribution
of shared functional variation under stabilizing selection}
\author{APR}
\date{\today}

@@

\onehalfspacing

@@

Genomic surveys of natural systems suggest that historical admixture among
diverged populations or closely related taxa commonly occors
\citep{brandvain2014speciation,skoglund2015ancient,suvorov2022widespread} and
is potentially widespread in primate
\citep{tung2017contribution,sorensen2023genome} and hominin
\citep{wolf2018outstanding,peter2020100} evolution. Admixture may therefore be
an important driver of phenotypic and molecular variation and shape the genetic
basis of complex traits. In humans, archaic introgression involving
Neanderthals and Denisovans has attracted considerable attention, with efforts
to describe the historical processes leading to observed distributions of
introgressed DNA in present-day populations
\citep{prufer2014complete,villanea2019multiple,chen2020identifying} and its
contribution to quantitative traits
\citep{sankararaman2016combined,wei2023lingering}.

Once introduced through admixture, introgressed alleles may be selected for or
against. Some introgressed haplotypes were positively selected in modern humans
\citep{huerta2014altitude,racimo2017signatures,enard2018evidence,gower2021detecting},
possibly due to locally adaptive variation that provided fitness advantages as
humans encountered novel environments. Despite some cases of adaptive
introgression, introduced alleles were more likely to have been selected
against in modern humans \citep{harris2016genetic,juric2016strength}. Since
Neanderthal and Denisovan population sizes were relatively small for hundreds
of thousands of years, theory predicts they would have accumulated deleterious
variants at an increased rate. Introgressed haplotypes loaded with more
deleterious mutations would then have quickly been selected against after
admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
humans shows a reduction of Neanderthal-related ancestry in enhancers and
regulatory regions
\citep{petr2019limits,telis2020selection,yermakovich2023long}. These
``deserts'' of Neanderthal ancestry support the hypothesis that introgressed
functional alleles were selected against
\citep{sankararaman2014genomic,sankararaman2016combined}.

There is growing genetic evidence that early \emph{H. sapiens} reciprocally
contributed to Neanderthal genomes
\citep{kuhlwilm2016ancient,hubisz2020mapping,harris2023diverse}. This
human-to-Neanderthal introgression would have occured tens or hundreds of
thousands of years prior to Neanderthal introgression in humans during modern
human global dispersal around 60 ka. This is supported by ``near modern''
\emph{H. sapiens} outside of African around 120--100ka or earlier
\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
with Neanderthals and providing opportunities for early contacts. While
estimates of the genomic contribution of early \emph{H. sapiens} to
Neanderthals vary, up to $\sim5\%$ of later Neanderthal genomes may have been
contributed through introgression from humans to Neanderthals
\citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
fewer deleterious alleles due to their larger long-term effective population
size, human-introgressed DNA would have been favored in Neanderthal genomes.
The replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports such a model of post-admixture selection in the Neanderthal
lineage \citep{posth2017deeply,petr2020evolutionary}.

Models for selection against introgressed alleles are often based on load
arguments (in particular, differences in the rate of accumulation of
unconditionally deleterious variation in populations of different sizes) or
hybrid incompatibilities \citep{muller1942isolating}. Such arguments, founded
in population genetics theory, rarely take into account selection operating on
phenotypic variation or the relationship between genetic and phenotypic
variation. Many phenotypic traits are thought to be under stabilizing selection
\citep{sanjak2018evidence,sella2019thinking}, including gene regulation
\citep{gilad2006natural,hodgins2015gene,price2022detecting}. Because some of
the strongest signals of selection against Neanderthal-introduced alleles are
in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
quantitative traits is particularly relevant to the dynamics of functional
genetic variation after introgression.

Stabilizing selection acts to maintain the phenotypic distribution of a trait
near some optimum, which is achieved by reducing phenotypic variation
(Figure~\ref{fig:one-pop}A). When the mean phenotype of the population is close
to the phenotypic optimum, classical models predict that alleles contributing
to such a trait are subject to underdominant selection, i.e., selection against
the minor allele at a locus \citep{robertson1956effect}. This has proven to be
a useful model for understanding allele frequencies of variants contributing to
a trait under stabilizing selection in single-population settings
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Stabilizing selection and additive genetic variance in one population.}
        \aprcomment{See comparisons to simulations assuming free recombination in
        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
        \aprcomment{$w$ and $f$ need to match text}
    }
    \label{fig:one-pop}
\end{figure}

In diverged populations, the genomic architecture of a trait under stabilizing
selection has a higher rate of turnover compared to neutrally evolving loci
\citep{yair2022population}. We might therefore expect an accumulation of fixed
differences in the \emph{H. sapiens} and Neanderthal lineages at loci
contributing to a trait, even when the mean phenotype in each population is
close to the same trait optimum. When a derived allele at high frequency in one
population is introduced to another population in which is was previously
absent, it will be at low frequency and subsequently selected against.
Likewise, if the ancestral allele is reintroduced to a population fixed for a
derived allele, the ancestral allele will be at low frequency and will be
selected against. In either case we should expect selection against the
introgressed allele, whether ancestral or derived and regardless the historical
relative sizes of the populations involved. This prediction contrasts with the
population-genetics ``load'' model, in which haplotypes with fewer deleterious
variants (such as those from the population with larger historical size) are
favored after introgression in either direction.

@@

We consider a polygenic trait such that an individual's (additive) genetic
value is the sum over all effects of alleles in their genome, so that for
individual $i$, \(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size
of the derived allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their
genotype at that locus (i.e., the number of alleles they carry at that locus).
For unlinked loci, the expected additive genetic variance is \(V_A = \sum_l
2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We ignore
dominance and epistatic effects, so that \(V_G = V_A\), assuming no linkage.
We further ingore environmental effects \citep{simons2018population}, so that
the phenotypic variance \(V_P=V_A\).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$ (typically set to zero), and we assume a Gaussian fitness function so
that relative fitness is given by \(f(G_i | O, V_S) = \exp{(-(G_i - O)^2 / 2
V_S)}\). $V_S$ can be interpreted as the strength of selection on the trait,
where larger $V_S$ implies weaker selection. For a population with mean
phenotype at (or very close to) the optimum, the mean fitness of the population
(assuming a roughly normal distribution of phenotypic values in the population)
is \[\bar{w} \approx \int_{-\infty}^\infty f(G | 0, V_S) \mathcal{N}(0, V_G) dG
= \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\] Thus, as the genetic variance
increases, mean fitness among individuals in the population decreases.

@@

If all alleles contribute equally to the trait
(with effect sizes \(\pm a\) occurring in equal proportion),
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_{G,t})\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known result
\[\tilde{V_G} \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
Mutations are not expected to each have the same effect size $a$, but
rather will be drawn from some distribution. Throughout, we assume mutation
effect sizes are drawn from a normal distribution with mean 0 and given
variance $V_M$.

\aprcomment{where does the $4\mu Ne V_M$ (Lande) result come from? point to fig 1}

@@

We extend the moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017inferring} to include underdominance with given
selection coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles
with effect size $a$ to the total genetic variance $V_G$ is then found by
computing the expected pairwise diversity from the SFS (with sample size $n$,
$\Phi_n$), as
\[V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da.\]
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
assuming a normal distribution of effect sizes for new mutations, the total
genetic variance is
\[V_G=\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration, and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, one remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{toy-admixture}A,B).

@@

\aprcomment{Start with ``Here we show that\dots''}
